---
title: "Untitled"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```
We load the libraries tidyverse and readxl, as well as janitor. First, read the data (both the metabolite concentration and the metabolic library).

```{r datareading}
metabolites <- readxl::read_xlsx('Assignment 1. Finding the gene that makes tomatoes pink.xlsx',
                         sheet = 'Metabolic data',
                         skip = 1,
                        .name_repair = make_clean_names)


metabolic_library <- readxl::read_xlsx('Assignment 1. Finding the gene that makes tomatoes pink.xlsx',
                         sheet = 'Metabolic Library',
                        .name_repair = make_clean_names)
```



Remove the first row, containing colors for the wildtype and ILs. Save this color data for possible later use. Convert the metabolite data into numeric if necessary.

```{r}
metabolites <- metabolites %>% 
  dplyr::slice(-1) %>% 
  mutate(across(wt_1:il01b_3, as.numeric),
         across(5:8, as.numeric))

```

```{r}
colordf <- tribble(
  ~type, ~colors,
  "wt1", "Red",
  "wt2", "Red",
  "wt3", "Red",
  "il01b_1", "Pink",
  "il01b_2", "Pink",
  "il01b_3", "Pink"
)
```

Compute averages and standard deviations of WT and IL01b
```{r}
metabolites <- metabolites %>% 
  rowwise() %>% 
  mutate(average_wt = mean(c(wt_1, wt_2, wt_3)),
         average_il01b = mean(c(il01b_1, il01b_2, il01b_3)),
         sd_wt = sd(c(wt_1, wt_2, wt_3)),
         sd_il01b = sd(c(il01b_1, il01b_2, il01b_3))) %>% 
  ungroup()

metabolites %>% 
  select(peaks, average_wt, sd_wt, average_il01b, sd_il01b) %>% 
  knitr::kable()
```
View the difference between the averages, and between the squared standard deviations (the variances).
```{r}
metabolites %>% 
  transmute(avg_diff = average_wt - average_il01b,
            variance_diff = sd_wt^2 - sd_il01b^2) %>% 
  knitr::kable()
```


We can see that 9 out of 12 rows have a higher WT average than their IL01b average.
We also see that the variances are not at all equal.
For safety, we can do a two-sided t-test on the data, between the WT and IL01b.
We check if the means are different, not if one is higher (or lower) than the other.

```{r}
ttestdata = NULL
for(i in 1:12) {
  x = as.vector(metabolites[i, 10:12])
  y = as.vector(metabolites[i, 13:15])
  singlettest = t.test(x, y, paired=FALSE, alternative = "two.sided")
  ttestdata[i] = singlettest$p.value
}
metabolites$ttest_wt_vs_il01b = ttestdata
```

We make a column with TRUE if the p-value of the t-test < 0.05, and FALSE if not.
```{r}
metabolites <- metabolites %>% 
  mutate(ttest_accepted = if_else(ttest_wt_vs_il01b < 0.05,
                                  TRUE, FALSE))
metabolites$ttest_accepted
```

So we see that the only one that's different, significantly, is the last one.
Now, calculate neutral masses for the metabolites.

```{r}
metabolites <- metabolites %>% 
  mutate(neutral_mass = negative_mass_m_h + 1.007825)

```

Find the metabolites with the closest retention time and closest mass. First, write a function to find the closest value.

```{r}
closest <- function(vec, value){
  vec[which(abs(vec-value) == min(abs(vec-value)))]
}
```

Then, calculate the closest retention times and masses.


```{r}
candidates_ret_time = NULL
ret_time_library = as.vector(metabolic_library$library_retention_time)
metabolite_names = as.vector(metabolic_library$metabolite_name)
metabolites$closest_ret_times = rep(0, 12)

for(i in 1:12){
  #retention time first
  closest_ret_time <- closest(ret_time_library, as.double(metabolites[i, 2]))
  if (length(closest_ret_time) == 1) {
    
    metabolites[i, 5] = closest_ret_time #retention time library
    metabolites[i, 7] = as.double(metabolites[i, 2]) - closest_ret_time  
    metabolites$closest_ret_times = closest_ret_time
    
  }else{
    closest_ret_time = closest_ret_time[1]
    print(paste('This one had multiple results: peak', i ))
    metabolites[i, 5] = closest_ret_time #retention time library
    metabolites[i, 7] = as.double(metabolites[i, 2]) - closest_ret_time
    metabolites$closest_ret_times = closest_ret_time
  }
  
 #retention time deviation
  ret_time_metabolite <- metabolic_library[ret_time_library == closest_ret_time, "metabolite_name"]
  
  candidates_ret_time[i] = ret_time_metabolite
}
```
For the masses:
```{r}
candidates_mass = NULL
mass_library = as.vector((metabolic_library$library_neutral_mass))
metabolites$closest_masses = rep(0, 12)

for(i in 1:12){  #Then, mass
  closest_mass <- closest(mass_library, as.double(metabolites[i, 4]))
  if(length(closest_mass) == 1){
      metabolites[i, 8] = ((metabolites[i, 4] - closest_mass)/closest_mass) * 1000000
      mass_metabolite <- metabolic_library[mass_library == closest_mass,
                                           "metabolite_name"]
      metabolites$closest_masses[i] = closest_mass
      
  }else{
    closest_mass = closest_mass[1]
    print(paste('This one had multiple results: peak', i ))
    
    metabolites[i, 8] = ((metabolites[i, 4] - closest_mass)/closest_mass) * 1000000
    mass_metabolite <- metabolic_library[mass_library == closest_mass,
                                         "metabolite_name"]
    metabolites$closest_masses[i] = closest_mass
  }


  candidates_mass[i] = mass_metabolite
}
```


Compare the lists of names. Use `View(met_names_compare)` to view the table, because the output is not displayed correctly in the console.
```{r, results=FALSE}
met_names_compare <- tibble(peaks = paste(rep(x= "Peak ", times=12), 1:12),
                            ret_time_names = candidates_ret_time,
                            mass_names = candidates_mass)
met_names_compare
```


Lets visualize the data into a barplot, showing the average metabolite concentration.
The error bars represent standard deviations.

```{r}

metabolites %>% 
  filter(peaks == "Peak 12") %>% 
  select(wt_1:il01b_3) %>% 
  pivot_longer(cols = everything(),
               names_to = "line",
               values_to = "concentration") %>% 
  mutate(type = as_factor(c(rep("WT", 3), rep("IL01b", 3)))) %>% 
  group_by(type) %>% 
  summarize(mean_concentration = mean(concentration),
            st_dev = sd(concentration)) %>% 
  ggplot(aes(x=type, y=mean_concentration))+
  geom_col(fill='azure3', color='black') +  
  labs(x = "", y = "Average concentration (Da)") +
  facet_wrap(~type, scales='free') +
  geom_errorbar(aes(ymin = (mean_concentration - st_dev), 
                    ymax = (mean_concentration + st_dev)),
                width = .5, position = 'dodge' ) + 
  theme_gray()
```
 

```{r, results=FALSE}
table1 <- metabolites %>% 
  select(1:5, 7, 8, closest_ret_times, closest_masses) %>% 
  left_join(met_names_compare, by='peaks',
            copy=TRUE,
            keep=FALSE)

table1$ret_time_names = met_names_compare$ret_time_names
table1$mass_names = met_names_compare$mass_names

table1
```
And the second one contains the results of the t-tests.
```{r}
table2 <- metabolites %>% 
  select(1, wt_1:ttest_accepted) %>% 
  select(!c(percent_sd_wt, percent_sd_il01b))

knitr::kable(table2, caption = "Table 2")  
```

