---
title: "Assignment 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(readxl)
library(Rtsne)
```

Load the libraries 'tidyverse', 'janitor', and 'readxl'. Then, load the data.
```{r datareading}
metabolic_data <- readxl::read_xlsx('Assignment 2. Analysis of the Solanum chmielewskii IL population.xlsx',
                                    sheet = 'Metabolic data W&A',
                                    skip = 7,
                                    .name_repair = make_clean_names)

```

We need to change the column names of some of the columns.

```{r}
names(metabolic_data)[1:8] <- c("peak", "scan_nr", "ret_min", "mass_mz", 
                                "sample_id", "annotation_level", "final_id",
                                "elemental_formula")


metabolic_data <- metabolic_data[-c(1, 2),] #remove the first two rows
```

Select only the first 25 metabolites.

```{r}
first_25 <- metabolic_data[1:25, ]
```

Make a scatterplot of quercetin rutinocide versus its isotope.

```{r}

quercetin <- paste(first_25[22, -(1:11)])
quercetin_isotope <- paste(first_25[23, -(1:11)])

df_quercetin <- tibble(quercetin = as.numeric(quercetin),
                      quercetin_isotope = as.numeric(quercetin_isotope))

ggplot(data=df_quercetin, aes(quercetin, quercetin_isotope)) +
  geom_point(alpha=0.5) +
  xlim(0, 8000) +
  ylim(0, 8000) +
  labs(x = "Quercetin values",
       y = "Quercetin isotope values")
  

```


Following code might not be used anymore, so i commented it out
```{r}
#UNNECCESSARY PROBABLY

# a_grouped <- first_25 %>% 
#   select(a_01a1:a_12d3) %>% 
#   pivot_longer(cols=everything(),
#                names_to = "individual_lines",
#                values_to = "expression") %>% 
#   mutate(il = stringr::str_extract(individual_lines, "^.{4}")) %>%
#   select(!individual_lines) %>%
#   pivot_wider(names_from = il,
#               values_from = expression) %>%
#   pivot_longer(cols=everything(),
#                names_to = "ID",
#                values_to = "value") %>% 
#   mutate(value = purrr::map(value, `length<-`, max(lengths(value)))) %>% #Magic line that adds NA to vectors to make them equal lengths
#   pivot_wider(names_from = "ID",
#               values_from = "value") %>% 
#   unnest(cols=everything())
```

Write a function to pivot a single metabolite row to a long format, for a single anova.


```{r}
longmaker <- function(data, colrange, rownumber){
  dplyr::select(data, {{colrange}}) %>% #double accolades for tidy selection
  dplyr::slice(rownumber) %>% 
  tidyr::pivot_longer(cols=everything(),
                      names_to = "individual_lines",
                      values_to = "expression") %>% 
  dplyr::mutate(il = dplyr::if_else(
    stringr::str_sub(individual_lines, start=3, end = 3) == "m", #IF
    stringr::str_extract(individual_lines, "^.{3}"), #if it's true
    stringr::str_extract(individual_lines, "^.{4}") #if it's false
    ))
}
```

a_01a1:a_12d3
Then, use the forloop to calculate the pvalues for each metabolite



```{r}
a_pvalue_results <- sapply(1:nrow(first_25), function(response){
  row_long <- longmaker(first_25, a_01a1:a_12d3, response)
  row_summary <-  anova(aov(expression ~ il, data = row_long))
  row_summary$`Pr(>F)`[1]
})

a_pvalue_results
```
Now do the same for the W column
```{r}
w_pvalue_results <- sapply(1:nrow(first_25), function(response){
  row_long <- longmaker(first_25, w_01a1:w_12d3, response)
  row_summary <-  anova(aov(expression ~ il, data = row_long))
  row_summary$`Pr(>F)`[1]
})

w_pvalue_results
```

Display the results inside a dataframe and do some corrections. You'll have to combine the pvalues first, in order for corrections to work out right. Now there are 50 pvalues, corrected once, instead of 2 times 25 pvalue corrections.

```{r}
pvalues_combined <- c(a_pvalue_results, w_pvalue_results)
pvalues_corrected <- p.adjust(pvalues_combined, method="holm")

metabolites_tested <- tibble(metabolites = first_25$final_id,
                             a_pvalues = a_pvalue_results,
                             w_pvalues = w_pvalue_results,
                             a_corrected = pvalues_corrected[1:25],
                             w_corrected = pvalues_corrected[26:50],
                             a_accepted = if_else(a_corrected < 0.05, TRUE, FALSE),
                             w_accepted = if_else(w_corrected < 0.05, TRUE, FALSE))

metabolites_tested
```

```{r}
metabolites_tested$metabolites[metabolites_tested$a_accepted == TRUE & metabolites_tested$w_accepted == TRUE]
```
```{r}
question5_df <- metabolites_tested %>%
  filter(a_pvalues <= 0.05 & w_pvalues <= 0.05) %>%
  mutate(combined = a_pvalues + w_pvalues) %>%
  arrange(combined) %>%
  select(metabolites, a_pvalues, w_pvalues)
question5_df
write.csv(question5_df, "question5df.csv")
```

```{r}
pivot_25 <- first_25 %>% 
  select(final_id, a_01a1:w_contr5) %>% 
  pivot_longer(cols = !final_id,
               names_to = "individual_lines",
               values_to = "expression") %>% 
  mutate(il = if_else(
    str_sub(individual_lines, start=3, end = 3) == "m", #IF
    str_extract(individual_lines, "^.{3}"), #if it's true
    str_extract(individual_lines, "^.{4}") #if it's false
    ))
```


```{r}

per_met_a_ttest <- sapply(1:25, function(i){
  row_long <- longmaker(first_25, 
                        colrange=a_01a1:a_m3,
                        rownumber = i)
  metabolite_m <- row_long$expression[row_long$il == "a_m"]
  
  result <- row_long %>% 
    filter(substr(il, 1, 1) == "a") %>% 
    group_by(il) %>% 
    summarize(ttest = t.test(expression, metabolite_m, 
                             alternative = "two.sided",
                             var.equal = FALSE)$`p.value`) %>% 
    select(ttest) %>% 
    slice(-13)
  result
})  

```


```{r}
per_met_w_ttest <- sapply(1:25, function(i){
  row_long <- longmaker(first_25, 
                        colrange=w_01a1:w_m3,
                        rownumber = i)
  metabolite_m <- row_long$expression[row_long$il == "w_m"]
  
  result <- row_long %>% 
    filter(substr(il, 1, 1) == "w") %>% 
    group_by(il) %>% 
    summarize(ttest = t.test(expression, metabolite_m, 
                             alternative = "two.sided",
                             var.equal = FALSE)$`p.value`) %>% 
    select(ttest) %>% 
    slice(-13)
  result
})  
```


```{r}
unlisted_a <- unlist(per_met_a_ttest, use.names = FALSE)
unlisted_w <- unlist(per_met_w_ttest, use.names = FALSE)

ttests_combined <- c(unlisted_a, unlisted_w)
ttests_corrected <- p.adjust(ttests_combined, method = "holm")

```



```{r}
question6_df <- tibble(metabolites = rep(first_25$final_id, each = 12),
                       line = rep(paste(rep("_", 12), 1:12), each=25) ,
                       a_combined = ttests_combined[1:300],
                       w_combined = ttests_combined[301:600]
                       )
```


```{r}
question6_df %>% 
  filter(a_combined < 0.05 & w_combined < 0.05) 

```
write.csv(question6_df, "question6df.csv")
Prep for pca.

```{r}
metabolite_columns <- pivot_25 %>% 
  pivot_wider(names_from = final_id,
              values_from = expression) %>% 
  select(-c(individual_lines, il))
```


```{r}
pca_res <- prcomp(metabolite_columns, center = TRUE, scale. = TRUE)
```


```{r}
pca_scores <- pca_res$x
```

```{r}
plot(pca_scores[,1:2], pch=21,    main="Metabolite data Scores")
```

```{r}
dist_eu <- philentropy::distance(metabolite_columns, method = "euclidean")
dist_eu_pca <- philentropy::distance(pca_scores[,1:2], method = "euclidean")
```

```{r}
d_data <- as.vector(dist_eu[upper.tri(dist_eu, diag = FALSE)])
d_pca <- as.vector(dist_eu_pca[upper.tri(dist_eu_pca, diag = FALSE)])

plot(d_data, d_pca, xlab = 'Distance in data space', ylab = 'Distance in PCA space' )
```

```{r}
ssw.data = NULL
ssw.pca = NULL
K = 20
for(k in 1 : K){
#km.res = kmeans(PCAscores[,1:2], k, iter.max = 25, nstart = 25)
#Clustering on original data
  km.res.data = kmeans(metabolite_columns, k, iter.max = 25, nstart = 25)
  ssw.data =  append(ssw.data, km.res.data$tot.withinss)
#Clustering on PCA scores
  km.res.pca = kmeans(pca_scores[,1:2], k, iter.max = 25, nstart = 25)
  ssw.pca =  append(ssw.pca, km.res.pca$tot.withinss)
}

```

```{r}
plot(1:K, ssw.data, xlab= 'Number of clusters', ylab = 'Total within SS',main = 'Clustering on original data space')
plot(1:K, ssw.pca, xlab= 'Number of clusters', ylab = 'Total within SS',main = 'Clustering on 2-D PCA space')

```



```{r}
km.res.pca = kmeans(pca_scores[,1:2], 4, iter.max = 25, nstart = 25)
cluster.id = km.res.pca$cluster
#Do pca 
pca.res <- prcomp(metabolite_columns, center = TRUE, scale. = TRUE)
PCAscores <- pca.res$x
#Plot PCA 
plot(PCAscores[,1:2], pch=21, col = cluster.id, main="Clusters from K-means clustering")

```




```{r}
res.tsne = Rtsne::Rtsne(metabolite_columns, perplexity = 1)

plot(res.tsne$Y, col = cluster.id)

```


```{r}
res.tsne = Rtsne::Rtsne(metabolite_columns, perplexity = 2)

plot(res.tsne$Y, col = cluster.id)
```




```{r}
res.tsne = Rtsne::Rtsne(metabolite_columns, perplexity = 5)

plot(res.tsne$Y, col = cluster.id)
```

```{r}
res.tsne = Rtsne::Rtsne(metabolite_columns, perplexity = 10)

plot(res.tsne$Y, col = cluster.id)
```


```{r}
res.tsne = Rtsne::Rtsne(metabolite_columns, perplexity = 25)

plot(res.tsne$Y, col = cluster.id)
```



```{r}

```
















